---
title: "Computing pairwise distances with fastDist"
author: "Vicente Castellar Sebastia"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Computing pairwise distances with fastDist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`fastDist` computes pairwise distances between the rows of two matrices `A`
and `B`. The single entry point is `fdist()`, which dispatches to C++
backends implemented with
[RcppArmadillo](https://cran.r-project.org/package=RcppArmadillo) and
parallelised with OpenMP where available.

Two design decisions distinguish `fastDist` from most alternatives:

* **Cross distances are first class.** Given an `m x k` matrix `A` and an
  `n x k` matrix `B`, `fdist(A, B, ...)` directly returns the `m x n` matrix
  whose entry `[i, j]` is the distance between row `i` of `A` and row `j` of
  `B`. Packages built around `stats::dist()` return square matrices over a
  single input, so obtaining `A x B` cross distances requires stacking the
  matrices and slicing the result, which wastes both time and memory.
* **One interface, many metrics.** The `method` argument selects among 19
  distance metrics, all sharing the same C++ infrastructure. The available
  backends are stored in a registry (`fdistregistry`) that can be inspected
  at run time.

```{r}
library(fastDist)
```

## Quick start

```{r}
set.seed(1)
A <- matrix(rnorm(5 * 3), nrow = 5, ncol = 3)
B <- matrix(rnorm(4 * 3), nrow = 4, ncol = 3)

# pairwise Euclidean distances between the rows of A and B
fdist(A, B, method = "euclidean")
```

Calling `fdist()` with a single matrix computes the full pairwise distance
matrix within its rows. The result is then a symmetric matrix with a zero
diagonal, and only the upper triangle is actually computed:

```{r}
fdist(A, method = "euclidean")
```

## Supported methods

`fdist()` selects the distance via the `method` argument. The available
methods are stored in `fdistregistry`:

```{r}
methods <- fdistregistry$get_entry_names()
descriptions <- vapply(
  methods,
  function(m) fdistregistry$get_entry(m)$description,
  character(1)
)
data.frame(method = methods, description = descriptions, row.names = NULL)
```

The remainder of this section groups the methods by the kind of data they
are intended for.

### General numeric data

The classical metrics `euclidean`, `manhattan`, `minkowski`, `supremum`
(Chebyshev) and `squared_euclidean` operate on arbitrary numeric data.

The `p` argument is only used by `method = "minkowski"` and sets the order
of the metric; it defaults to `p = 2`. `p = 1` recovers the Manhattan
distance and `p = 2` the Euclidean distance:

```{r}
fdist(A, B, method = "minkowski", p = 1)
fdist(A, B, method = "manhattan")
```

`standardized_euclidean` scales each feature by its sample variance
(estimated from `A`), which makes the metric insensitive to the units in
which each column is measured:

```{r}
fdist(A, B, method = "standardized_euclidean")
```

### Shape-based metrics: correlation, cosine and Spearman

These metrics compare the *profile* of two rows rather than their magnitude.
They are popular for expression data, text embeddings and time series
profiles:

```{r}
fdist(A, B, method = "correlation")
fdist(A, B, method = "cosine")
fdist(A, B, method = "spearman")
```

`correlation` is one minus the Pearson correlation, `spearman` is one minus
Spearman's rank correlation and `cosine` is one minus the cosine similarity.

### Non-negative / compositional data

The methods `bray_curtis`, `hellinger`, `chi_squared` and `jensen_shannon`
are intended for non-negative data such as histograms, species abundances or
probability distributions. For `jensen_shannon` each row is normalised to
sum to one before the computation:

```{r}
P <- matrix(runif(5 * 4, 0.1, 5), nrow = 5)
Q <- matrix(runif(3 * 4, 0.1, 5), nrow = 3)
fdist(P, Q, method = "jensen_shannon")
fdist(P, Q, method = "bray_curtis")
```

### Binary and categorical data: Hamming and Jaccard

`hamming` counts the number of positions in which two rows differ, which
makes it appropriate for categorical codes or binary indicators:

```{r}
X <- matrix(rbinom(6 * 8, 1, 0.5), nrow = 6)
fdist(X, method = "hamming")
```

`jaccard` treats each row as a set (non-zero entries are members) and
returns one minus the ratio of the intersection to the union:

```{r}
fdist(X, method = "jaccard")
```

### Mixed-scale data: Gower

`gower` rescales every feature by its range (computed from `A`) before
averaging the absolute differences, so features measured on very different
scales contribute equally:

```{r}
mixed <- cbind(
  age    = c(25, 47, 31, 60),
  income = c(20000, 52000, 34000, 71000),
  score  = c(0.3, 0.9, 0.5, 0.7)
)
fdist(mixed, method = "gower")
```

### Geographic data: Haversine

The `haversine` method computes great-circle distances in kilometres between
points given as `(latitude, longitude)` in degrees. The input must have
exactly two columns:

```{r}
cities <- matrix(
  c(40.4168, -3.7038,   # Madrid
    41.3851,  2.1734,   # Barcelona
    48.8566,  2.3522),  # Paris
  ncol = 2, byrow = TRUE
)
fdist(cities, method = "haversine")
```

### Covariance-aware data: Mahalanobis

`method = "mahalanobis"` always works within the rows of `A` (any `B` is
ignored) and uses the sample covariance matrix estimated from `A`:

```{r}
fdist(A, method = "mahalanobis")
```

## Performance

All backends are written in C++ (RcppArmadillo) and parallelised with OpenMP.
For symmetric problems (`B = NULL`) only the upper triangle is computed, and
for the Euclidean family the computation is reformulated as a matrix product
plus row norms, which is substantially faster than the naive double loop.

The package sources ship two reproducible benchmark scripts (they are not
installed with the package; see the `benchmark/` directory of the GitHub
repository):

* `benchmark/benchmark_parallelDist.R` compares `fdist()` against
  `parallelDist::parDist()`.
* `benchmark/benchmark_comparison.R` compares `fdist()` against
  `parallelDist::parDist()`, `proxy::dist()` and `rdist::cdist()` for every
  metric the four packages share, after checking that all packages agree on
  the result.

A typical invocation looks like this:

```{r, eval = FALSE}
source("benchmark/benchmark_comparison.R")
results <- run_benchmark(
  a_rows = 1000,
  b_rows = c(1000, 5000, 10000),
  n_features = 1000
)
print(results)
plot_benchmark(results)   # requires ggplot2
```

As a reference, the table below reports the mean time in seconds to compute
the cross-distance matrix between `A` (`1000 x 1000`) and `B` (with a growing
number of rows), measured with `microbenchmark` on an Intel i5 10400 (6
cores). `parallelDist` computes square matrices only, so its timings include
the necessary `rbind()` + slicing step described above:

| method    | rows of B | fastDist (s) | parallelDist (s) |
|-----------|-----------|--------------|------------------|
| euclidean | 1000      | 3.08         | 5.13             |
| euclidean | 10000     | 31.66        | 51.90            |
| euclidean | 20000     | 79.66        | 99.47            |
| manhattan | 1000      | 2.30         | 4.68             |
| manhattan | 10000     | 38.46        | 54.81            |
| manhattan | 20000     | 89.92        | 106.93           |
| minkowski | 1000      | 4.31         | 9.19             |
| minkowski | 10000     | 64.18        | 94.07            |
| minkowski | 20000     | 139.66       | 185.13           |

Actual timings depend on hardware, BLAS library and number of OpenMP
threads; run the scripts on your own machine for meaningful numbers.

## Session information

```{r}
sessionInfo()
```
