---
title: "Computing pairwise distances with fastDist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing pairwise distances with fastDist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`fastDist` computes pairwise distances between the rows of two matrices `A`
and `B`. The single entry point is `fdist()`, which dispatches to C++ backends
implemented with [RcppArmadillo](https://cran.r-project.org/package=RcppArmadillo)
and parallelised with OpenMP where available.

Given an `m x k` matrix `A` and an `n x k` matrix `B`, `fdist()` returns an
`m x n` matrix whose entry `[i, j]` is the distance between row `i` of `A` and
row `j` of `B`. When `B` is omitted, distances are computed within the rows of
`A` (the result is then a symmetric `m x m` matrix with a zero diagonal).

```{r}
library(fastDist)
```

## A first example

```{r}
set.seed(1)
A <- matrix(rnorm(5 * 3), nrow = 5, ncol = 3)
B <- matrix(rnorm(4 * 3), nrow = 4, ncol = 3)

# pairwise Euclidean distances between the rows of A and B
fdist(A, B, method = "euclidean")
```

Calling `fdist()` with a single matrix computes the full pairwise distance
matrix within its rows:

```{r}
fdist(A, method = "euclidean")
```

## Supported methods

`fdist()` selects the distance via the `method` argument. The available
methods are:

```{r}
fdistregistry$get_entry_names()
```

A short description of each is stored in the registry:

```{r}
methods <- fdistregistry$get_entry_names()
descriptions <- vapply(
  methods,
  function(m) fdistregistry$get_entry(m)$description,
  character(1)
)
data.frame(method = methods, description = descriptions, row.names = NULL)
```

### Minkowski distance and the `p` argument

The `p` argument is only used by `method = "minkowski"` and sets the order of
the metric. `p = 1` recovers the Manhattan distance and `p = 2` the Euclidean
distance:

```{r}
fdist(A, B, method = "minkowski", p = 1)
fdist(A, B, method = "manhattan")
```

### Distances for non-negative / compositional data

Some methods (`bray_curtis`, `hellinger`, `chi_squared`, `jensen_shannon`)
are intended for non-negative data such as histograms, abundances or
distributions:

```{r}
P <- matrix(runif(5 * 4, 0.1, 5), nrow = 5)
Q <- matrix(runif(3 * 4, 0.1, 5), nrow = 3)
fdist(P, Q, method = "jensen_shannon")
```

### Haversine distance

The `haversine` method computes great-circle distances in kilometres between
points given as `(latitude, longitude)` in degrees. The input must have
exactly two columns:

```{r}
cities <- matrix(
  c(40.4168, -3.7038,   # Madrid
    41.3851,  2.1734,   # Barcelona
    48.8566,  2.3522),  # Paris
  ncol = 2, byrow = TRUE
)
fdist(cities, method = "haversine")
```

### Mahalanobis distance

`method = "mahalanobis"` always works within the rows of `A` (any `B` is
ignored) and uses the sample covariance matrix estimated from `A`:

```{r}
fdist(A, method = "mahalanobis")
```

## Performance

`fastDist` is designed to be fast for large matrices. A reproducible
benchmark comparing `fdist()` against `parallelDist::parDist()` for the
Euclidean, Manhattan and Minkowski methods is available in
`benchmark/benchmark_parallelDist.R` in the package sources. See the package
README for an example of the results.

## Session information

```{r}
sessionInfo()
```
